---
title: Development status
package: anndataR
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Development status}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This vignette provides an overview of the current development status of the `r Biocpkg("anndataR")` package.
It provides details on the current implementation of different features as well as listing known issues.

```{r impl-table, include=FALSE}
library(tibble)
library(knitr)
library(rprojroot)
library(stringr)
library(purrr)
library(dplyr)
library(tidyr)

proj_root <- find_root(has_file("DESCRIPTION"))
source <- list.files(proj_root, pattern = "*.R$", recursive = TRUE)

# look for trackstatus comments
status_lines <- map_df(source, function(path) {
  lines <- readLines(paste0(proj_root, "/", path))
  line_numbers <- grep("# trackstatus:", lines)

  map_df(line_numbers, function(line_number) {
    tryCatch(
      {
        line_stripped <- gsub(" *# trackstatus: *", "", lines[[line_number]])
        line_split <- str_split_1(line_stripped, ", *")
        vals_split <- str_split(line_split, " *= *")
        names <- vapply(vals_split, function(x) x[[1]], character(1))
        values <- lapply(vals_split, function(x) x[[2]])
        df <- data.frame(setNames(values, names), check.names = FALSE)
        df$source_file <- path
        df$line_number <- line_number
        df
      },
      error = function(e) {
        message("Error in ", path, " at line ", line_number, ": ", e$message)
        NULL
      }
    )
  })
})

# split feature id into prefix and slot columns
strip_prefix <- c(
  "get_" = "Getter",
  "test_get_" = "Getter test",
  "set_" = "Setter",
  "test_set_" = "Setter test"
)
status_lines_proc <- status_lines %>%
  mutate(
    prefix = str_extract(feature, "^(test_)?[gs]et_"),
    slot = str_replace(feature, "^(test_)?[gs]et_", "")
  ) |>
  select(-feature)

# combine with missing fields
status_lines_required <- crossing(
  class = c("InMemoryAnnData", "HDF5AnnData", "Seurat", "SingleCellExperiment"),
  prefix = c("get_", "test_get_", "set_", "test_set_"),
  slot = c(
    "X",
    "layers",
    "obs",
    "var",
    "obs_names",
    "var_names",
    "obsm",
    "varm",
    "obsp",
    "varp",
    "uns",
    "raw"
  ),
  status = "missing"
)
status_required <- bind_rows(
  status_lines_proc,
  anti_join(
    status_lines_required,
    status_lines_proc,
    by = c("class", "prefix", "slot")
  )
)

# Fill in non-required slots
status <- status_required |>
  group_by(class) |>
  complete(prefix, slot, fill = list(status = "missing")) |>
  ungroup()

# check duplicated status
status_dup <- status |>
  group_by(class, prefix, slot) |>
  filter(n() > 1) |>
  ungroup() |>
  mutate(
    str = paste0(
      source_file, "#", line_number, " -- ", class, " ", prefix, slot
    )
  )
if (nrow(status_dup) > 0) {
  stop(
    "Duplicated status lines found:\n", paste(status_dup$str, collapse = "\n")
  )
}

# create formatted data frame for printing
status_formatted <- status |>
  mutate(
    prefix_formatted = strip_prefix[prefix],
    status_formatted = ifelse(
      status == "missing",
      "",
      paste0(
        "[", c("done" = "✅", "wip" = "🚧", "missing" = "")[status],
        "](https://github.com/scverse/anndataR/blob/main/", source_file,
        "#L", line_number, ")"
      )
    )
  ) |>
  select(class, Slot = slot, prefix_formatted, status_formatted) |>
  spread(prefix_formatted, status_formatted)
```

# Objects

These tables show the status of the implementation of different `AnnData` back ends.

```{r impl-table-objects, echo=FALSE, results="asis"}
# loop over each of the classes and print the table as markdown
all_classes <- unique(status_formatted$class)
object_classes <- all_classes[!all_classes %in% c("Seurat", "SingleCellExperiment")]

for (class_name in object_classes) {
  cat("## `", class_name, "`\n\n", sep = "")

  df <- status_formatted |>
    filter(class == class_name) |>
    select(-class) |>
    knitr::kable(escape = FALSE, align = "lcccc")

  cat(paste(as.character(df), collapse = "\n"))
  cat("\n\n")
}
```

# Conversion

These tables show the implementation status of conversion between `AnnData` and other objects.

```{r impl-table-conversion, echo=FALSE, results="asis"}
# loop over each of the classes and print the table as markdown
for (class_name in c("SingleCellExperiment", "Seurat")) {
  cat("## `", class_name, "`\n\n", sep = "")

  df <- status_formatted |>
    filter(class == class_name) |>
    select(
      Slot,
      From = Setter,
      `From test` = `Setter test`,
      To = Getter,
      `To test` = `Getter test`
    ) |>
    knitr::kable(escape = FALSE, align = "lcccc")

  cat(paste(as.character(df), collapse = "\n"))
  cat("\n\n")
}
```

# Known issues

This section lists current known issues in `r Biocpkg("anndataR")`.
Only certain types of issues are listed here, for additional issues see the [GitHub issue tracker](https://github.com/scverse/anndataR/issues?q=sort%3Aupdated-desc+is%3Aissue+is%3Aopen).

```{r known-issues, echo=FALSE, results="asis"}
known_issues_path <- system.file("known_issues.yaml", package = "anndataR")
data <- yaml::read_yaml(known_issues_path)$known_issues

for (i in seq_along(data)) {
  str <- paste0(
    "## Issue: ", data[[i]]$description, "\n\n",
    " * Affected backend: ", paste0("`", data[[i]]$backend, "`", collapse = ", "), "\n",
    " * Affected slot(s): ", paste0("`", data[[i]]$slot, "`", collapse = ", "), "\n",
    " * Affected dtype(s): ", paste0("`", data[[i]]$dtype, "`", collapse = ", "), "\n",
    " * Probable cause: ", data[[i]]$process, "\n",
    " * To investigate: ", data[[i]]$to_investigate, "\n",
    " * To fix: ", data[[i]]$to_fix, "\n\n",
    "### Error message\n\n",
    paste(paste0("    ", strsplit(data[[i]]$error_message, "\n")[[1]], "\n"), collapse = ""), "\n\n",
    "### Proposed solution\n\n",
    data[[i]]$proposed_solution, "\n\n"
  )
  cat(str)
}
```

# Session info

```{r sessioninfo}
sessionInfo()
```
